---
title: sentinel
output: github_document
---

S3 class that allows different flavors of missing in numeric vectors.

One can divide measures into two groups: qualitative and quantitative. However,
record formats often mix the two. Some of the values are simply interpreted as
is: a `2` is a 2. Some of the values are codes which represent qualities instead
of numbers: an `8` means the measure's not applicable. These are sometimes
called "sentinel values." And, of course, some values are just plain missing.

When handling these data in R, a common idiom is to split the column in twain: a
numeric vector for the quantitative and a factor for the qualitative. This is
the simplest solution and will often work fine. But it does something risky: it
separates linked data. The user must remember to keep them together, and usually
does this with clever variable or column names.

Clever is bad. Code with `my_data[, paste0(vars, c("_num", "_flag"))]` is hard
to read. Code with `get` is hard to follow.

The `sentinel` package offers the `sentineled` class to bundle numeric and
categorical missing values into a single object.

```{r sentineled}
library(sentinel)

x <- sentineled(
  c(10, 20, 98, 99, NA),
  sentinels = c(98, 99),
  labels    = c("refused", "not recorded")
)
x
```

The numbers are numbers, the categories are categorical, and the unknowns are
just unknown.

## Still a vector

A `sentineled` object is a vector. When subsetting, a it will remain a
`sentineled` object with the same possible sentinel values.

```{r subsetting}
x[1]
x[1:2]
x[[3]]
x[x < 15]
```

A `sentineled` vector can be used in arithmetic, with all non-missing values
acting like normal numeric values. If possible, a `sentineled` object with the
appropriate sentinel values will be the result.

```{r sentineledArithmetic}
mean(x, na.rm = TRUE)
x / 100
```

## Using the missing values

The sentinel codes are treated as missing, but the different categories of
missing are stored as a factor vector in the `"sentinels"` attribute of the
object. Use the `sentinels` function to access them.

```{r sentinels}
sentinels(x)
x[sentinels(x) != "refused"]
```

Notice that, for the non-missing values in `x`, their respective sentinel codes
are blanks (`""`).

```{r sentinelsCharacter}
as.character(sentinels(x))
```

It's recommended to use explanatory sentinel levels for all expected types of
missing. That way, if a value is shown as just plain `NA`, it's a sign something
went wrong in the analysis.
